---
title: "Workshop #6: Mapping injustice"
author: "ENST 320"
date: "Spring 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
print_key = FALSE
```

**Instructions:**  

1. Save a copy of this document to your own Drive so that you can edit it.   

2. You may work on this with others, but your answers must be your own, typed manually by you.  

3. For each question, replace the "*Your code/answer/plot goes here*" line with your response. (For plots, you can take a screenshot of your plot from `R`).  

4. When you are done, save your doc as a PDF then submit via this [form](https://forms.gle/2QrE6vTmWwup4DMv6).  

Note that submissions will be assessed using [this rubric](https://docs.google.com/presentation/d/1PesPBMLaJqNBtc4mFJ8ymcsZaT1UjWAXQZyi10yZY84/edit#slide=id.g1c31560e1bb_0_7).  

&nbsp;  

#### (1) Your name:  
**Your answer goes here.**
&nbsp;  
&nbsp;  

#### (2) "Library" packages
*Show the code needed to load the following packages for use in your R session. Make sure these lines of code work when you run them. If any of these packages are not installed on your R account, install them and show the code for how you did so.*

- `dplyr`
- `ggplot2`
- `leaflet`
- `sf`
- `viridis`

**Your answer goes here.**

&nbsp;  

```{r , echo=print_key, eval=TRUE, include=print_key, message=FALSE, warning=FALSE}
# Answer key
library(dplyr)
library(ggplot2)
library(leaflet)
library(sf)
library(viridis)
```

&nbsp;  

#### The SEPHER dataset

Our goal today is to explore and map the relationship between social and environmental injustice. To do so we will draw upon Columbia University's famous [SEPHER dataset](https://www.ciesin.columbia.edu/data/sepher/) (see related publication [here](https://www.liebertpub.com/doi/10.1089/env.2021.0059)). This dataset summarizes socioeconomic demographics, housing status, and exposure to environmental risks (e.g., tornadoes, wildfires, flooding, etc.) for every census tract in the United States (see maps of those tracts [here](https://www.google.com/search?q=map+of+census+tracts&rlz=1C5GCEM_enUS914US914&sxsrf=AJOqlzVuq8FlethjplKv-DWdei7KlSb9Xg:1678462715520&source=lnms&tbm=isch&sa=X&ved=2ahUKEwimirXv2NH9AhWiH0QIHfnpD0kQ_AUoAnoECAEQBA&biw=1792&bih=849&dpr=2#imgrc=IqwlVkgAfwox8M)).

This is dataset is *enormous*: for each census tract (there are about 86,000), the SEPHER dataset contains over 1,600 variables for *every year* from 2000 to 2016. It's a goldmine!

To make this dataset more manageable, Ezell has filtered it down to only 174 variables in the year 2016 only. You can load that filtered dataset with the following code: 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
load(url('https://github.com/ericmkeen/sewanee_esus/blob/master/08_mapping_injustice/SEPHER_2016.rds?raw=true'))
```

Check out your "Environment", and you should see a new data object named `sepher`. Here's what the first six rows of the first 10 columns should look like: 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
sepher[1:6, 1:10]
```

Every row is a census tract. To understand what each column means, `load` & `View` this key: 

```{r , echo=TRUE, eval=print_key, include=TRUE, message=FALSE, warning=FALSE}
load(url('https://github.com/ericmkeen/sewanee_esus/blob/master/08_mapping_injustice/key_sepher.rds?raw=true'))

# That code saves a dataframe named `key`.
key %>% View

```

This dataset allows you to explore the correlation between socioeconomic factors, housing status, and environmental risk across the United States.

#### (3) Produce a scatterplot displaying the relationship between the percent-whiteness of a community (column `PCT.WHITE_2016`) and its poverty rate (column `POVERTY.RATE_2016`).  
 
&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;  

```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
# Answer key
ggplot(sepher, aes(y=POVERTY.RATE_2016,
                   x=PCT.WHITE_2016)) +
  geom_point(alpha=.1) + 
  labs(title = 'More whiteness, less poverty')
```

&nbsp;  

#### (4) Produce a plot comparing the distribution of environmental risk exposure (`RISK_SCORE`) across census tracts in California (CA) vs. Florida (FL). Note that state abbreviations are in the column `ST_ABBR`. 

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;

```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
# Answer key
ggplot(sepher %>% filter(ST_ABBR %in% c('CA', 'FL')),
       aes(x=RISK_SCORE,
           fill=ST_ABBR)) +
  geom_density(alpha=.4)
```
&nbsp;

#### (5) Produce a plot that explores the relationship between environmental risk exposure (`RISK_SCORE`) and median household income (`MEDIAN.HOUSEHOLD.INCOME_2016`).  

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;

```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
# Answer key
ggplot(sepher, aes(y=RISK_SCORE,
                   x=MEDIAN.HOUSEHOLD.INCOME_2016)) +
  geom_point(alpha=.1)
```
&nbsp;

#### (6) Devise your own justice-related question and make a plot that answers it
  

&nbsp;  
**Your question goes here.**   
&nbsp;  

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;


&nbsp;  


## Mapping census tracts

Here is a dataset with census tract polygons for mapping. 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
load(url('https://github.com/ericmkeen/sewanee_esus/blob/master/08_mapping_injustice/census_tracts.rds?raw=true'))
```

After running that code, your Environment should have a dataset named `tracts`.  That dataset is known as a `sf` object, which refers to the package (`sf`) that is used to work with spatial data in `R`.  Check out the head of the dataset:  

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
tracts %>% head
```

That final column, `geometry`, contains all the data that `R` needs to know in order to add polygons to a map.  But you can still treat this as a `dataframe` that you can `filter`, `mutate`, etc. 

To learn how to map, let's filter the data to just a single state, since trying to plot the entire dataset at once might crash your computer:  

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
cali <- tracts %>% filter(STATE == 'CALIFORNIA')
```

Now add these California census polygons a `ggplot()` using a new geom, `geom_sf()`: 

```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
ggplot(cali) +
  geom_sf()
```

&nbsp;  

## Join SEPHER and census tract data together

Use the following code to join the `sepher` and `tracts` datasets together: 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
mr <- left_join(tracts, sepher, by='FIPS')
```

Now you have census tract polygon details associated with all of your  socioeconomic/environmental data. 

&nbsp;  


## Make a "chloropleth"

A chloropleth is a map in which polygons are color-coded according to data. 

To make your first chloropleth, let's filter our dataset down to Tennessee: 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
mr_tn <- mr %>% filter(STATE == 'TENNESSEE')
```

Now let's color-code Tennessee's census tracts by their per-capita income: 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ggplot(mr_tn,
       aes(fill=E_PCI_2016)) +
  geom_sf(color=NA) +
  labs(title='Per-capita income by census tract')
```

Note that we added `color = NA` to the `geom_sf()` call to get rid of thick borders around the census tracts.  

These tracts are being color-coded according to a default color scale in `ggplot()`, but you can specify other color scales thanks to the [`viridis` package](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html): 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ggplot(mr_tn,
       aes(fill=E_PCI_2016)) +
  geom_sf(color=NA) +
  scale_fill_viridis(option='magma')
```


#### (7) Create a chloropleth map of a state of your own choice, color-coded according to an environmental risk variable of your choice. 

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;

```{r , echo=print_key, eval=print_key, include=TRUE, message=FALSE, warning=FALSE}
# Example answer
mr_cali <- mr %>% filter(STATE == 'CALIFORNIA')

ggplot(mr_cali, aes(fill=RISK_SCORE)) +
  geom_sf(color=NA) +
  labs(title='FEMA risk score')

```

&nbsp;

## Manually define your map range

You can also crop the census tract data to a range of your choice. For example, you could use this code to crop data to show all census tracts within 1/2 a degree of Sewanee, 

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
# Sewanee's LAT/LON
x <-  -85.92
y <- 35.2

# Crop your tracts
socu <- st_crop(mr,
                xmin = x - 0.5,
                ymin = y - 0.5,
                xmax = x + 0.5,
                ymax = y + 0.5)
```

```{r , echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ggplot(socu,
       aes(fill=POPULATION)) +
  geom_sf() +
  geom_point(aes(x=x, y=y), 
             color='orange', 
             size=3)
```

#### (8) Create a chloropleth map of census tracts in and around the U.S. city of your own choice. 

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;

&nbsp;


## Air quality

Ezell also downloaded a dataset from the EPA that estimates cancer risk for each census tract based upon air quality. Load that dataset with this code:  

```{r , echo=TRUE, eval=print_key, include=TRUE, message=FALSE, warning=FALSE}
load(url('https://github.com/ericmkeen/sewanee_esus/blob/master/08_mapping_injustice/cancer_risk.rds?raw=true'))
```

Your environment should now have an object named `air`.  

Join the `air` dataset to your `mr` dataset. 

```{r , echo=TRUE, eval=print_key, include=TRUE, message=FALSE, warning=FALSE}
mra <- left_join(mr, air, by='FIPS')
```

#### (9) Produce a plot that displays the relationship between a census tract's whiteness (`PCT.WHITE_2016`) and its cancer risk (`cancer_risk`).  Set the y-axis range to 0-70.
 
&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;  


```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
ggplot(mra,
       aes(x=PCT.WHITE_2016,
           y=cancer_risk)) +
  ylim(0, 70) +
  geom_point(alpha=.1)
```

&nbsp;  

Let's explore those variables in a different way. Create a new column that categorizes each census tract as `very white` or `less white` based on a `PCT.WHITE_2016` threshold of 0.5. 

You can do this pretty easily using an `ifelse` statement. Here's how it works:  

```{r , echo=TRUE, eval=print_key, include=TRUE, message=FALSE, warning=FALSE}
mra <- mra %>% 
  mutate(whiteness = ifelse(PCT.WHITE_2016 > 0.5, 
                            'more white', 
                            'less white'))
```

#### (10) Before we plot anything, filter the dataset to rows where `whiteness` is not `NA` and `cancer_risk` is less than 100.  

&nbsp;  
**Your code goes here.**   
&nbsp;  


```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
# Answer key
mra <- mra %>% filter(is.na(whiteness) == FALSE, cancer_risk < 100)
```

&nbsp;  


#### (11) Now compare the two whiteness categories using a boxplot (use `geom_boxplot()`).

&nbsp;  
**Your code goes here.**   
&nbsp;  

&nbsp;  
**Your screenshot goes here.**   
&nbsp;

```{r , echo=print_key, eval=print_key, include=print_key, message=FALSE, warning=FALSE}
# Answer key
ggplot(mra,
       aes(x=whiteness, y=cancer_risk)) +
  geom_boxplot()
```

&nbsp; 



