---
title: "Workshop #1: Greenhouse gas emissions"
author: "ENST 320"
date: "Spring 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Instructions:**  

1. Save a copy of this document to your own Drive so that you can edit it.   

2. You may work on this with others, but your answers must be your own, typed manually by you.  

3. For each question, replace the "*Your code/answer/plot goes here*" line with your response. (For plots, you can take a screenshot of your plot from `R`).  

4. When you are done, save your doc as a PDF then submit via this [form](https://forms.gle/2QrE6vTmWwup4DMv6).  

Note that submissions will be assessed using [this rubric](https://docs.google.com/presentation/d/1PesPBMLaJqNBtc4mFJ8ymcsZaT1UjWAXQZyi10yZY84/edit#slide=id.g1c31560e1bb_0_7).  

&nbsp;  

#### (1) Your name:  
**Your answer goes here.**
&nbsp;  
&nbsp;  

#### (2) "Library" packages
*Show the code needed to load the following packages for use in your R session. Make sure these lines of code work in your `R` session.*

- `dplyr`
- `ggplot2`
- `readr`

**Your answer goes here.**
&nbsp;  
&nbsp;  


```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

#### (3) "Read in" the data
*In this workshop we will use a massive carbon emissions database maintained by the website "Our World in Data". The URL for this dataset is:*

`https://nyc3.digitaloceanspaces.com/owid-public/data/co2/owid-co2-data.csv`  

*Show the code needed to save this link into an object named `url` (make sure the link is in quotes), then read the data in the link into R using the function `read_csv()`, saving the data into a new object named `carbon`.*

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (4) View your data
*Show the code you would use to view this entire dataset within `R`.*  

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (5) Get column names
*Show the code to view just the column names of the dataset.*  

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (6) Select certain columns
*Show the code to select only a few of the dataset's columns, saving the subset version of the data into a new object named `subcarbon`. The columns to keep are* `country`, `year`, `population`, `gdp`, `co2`, `cumulative_co2`, *and* `co2_per_gdp`.   

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  


## Historical trends in carbon emissions from the United States

#### (7) Filter the dataset
*Show the code to filter `subcarbon` to show only rows where `country` is equal to `"United States"`. Store this filtered dataset in a new object named `usa`.*  

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (8) Annual CO2e emissions by the United States
*Show the code to produce a beautiful line graph of annual trends in CO2e emissions by the United States. Then provide a screenshot of your plot.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (9) Cumulative CO2e emissions by the United States
*Show the code to produce a beautiful line graph of cumulative CO2e emissions by the United States. Then provide a screenshot of your plot.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp; 

#### (10) Annual CO2e emissions *per capita* in the United States
*Show the code to produce a beautiful line graph of annual trends in per-capita CO2e emissions by the United States. Then provide a screenshot of your plot.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (11) The U.S. "drawdown"
*Based on your graphs, in what year did United States emissions peak?*  

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (12) Explain emissions types
*Explain the difference between annual emissions, cumulative emissions, and per-capita emissions.*  

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

## Comparing the top two emitters: United States & China.

Re-filter `subcarbon` to include data for both the United States AND China, and save the result in a new object named `top_two`. Filtering to multiple countries is a bit different than filtering to a single country, so here is the code you need to do so:

```{r , eval=FALSE}
top_two <- subcarbon %>% filter(country %in% c('United States', 'China'))
```

#### (13) Annual trends
*Show the code to produce a beautiful line graph of annual emissions trends in these two countries. To do so, you will need to make sure `ggplot()` knows that line `color` should be deteremined by the `country`. Then provide a screenshot of your plot.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (14) Cumulative trends
*Show the code and a screenshot of your plot for comparing cumulative emissions in the two countries over time.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (15) Per-capita trends
*Show the code and a screenshot of your plot for comparing per-capita emissions in the two countries over time.*  

&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (16) Which country currently emits the most CO2 annually? 

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  

#### (17) Which country currently emits the most on a per-capita basis?

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp; 


#### (18) Which country has emitted the most cumulatively over time?

&nbsp;  
**Your answer goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp; 


#### (19) Design your own plot!
*Use the `subcarbon` dataset to produce a plot of your own focus and design. It can use whichever columns you wish, and as many countries as you wish. Make your plot beautiful, and give it a title to make it clear what you are showing.*

*Recall that you can add a title to your plot using the line...* 

`+ labs(title = "My title is here")`  


&nbsp;  
**Your code goes here.**    

**Your screenshot goes here.**   

```{r , echo=FALSE, eval=FALSE}
# Answer key goes here
```

&nbsp; 
&nbsp;  



