---
title: 'Geographies of longevity and wealth'
author: "`R` puzzle #5: Environment & Sustainability (2022)"
date: "2/18/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Answer these questions in a separate script.*

*** 

### Setup

- First, install the `gapminder` package.   

```{r, echo=TRUE, eval=FALSE}
install.packages('gapminder')
```

- Load this and other packages:  

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(gapminder)
library(ggplot2)
library(dplyr)
```

- Load data from the `gapminder` package:  

```{r, echo=TRUE, eval=TRUE}
data(gapminder)
```

*** 

### Part 1

**Write code that answers these questions:** 

**1A.** Look at the top six rows of this dataset:

```{r, echo=FALSE, eval=FALSE}
head(gapminder)
```

**1B.** How many countries are represented in this dataset?

```{r, echo=FALSE, eval=FALSE}
gapminder$country %>% unique %>% length
```

**1C.** What unique years are represented?

```{r, echo=FALSE, eval=FALSE}
gapminder$year %>% unique
```

**1D.** What was the total world population in 1952?

```{r, echo=FALSE, eval=FALSE}
gapminder %>% filter(year==1952) %>% summarize(pop = sum(pop))
```

**1E.** In 2007? *(This time, provide your answer in billions, e.g., `3.2` instead of `3,200,000,000`)*

```{r, echo=FALSE, eval=FALSE}
gapminder %>% filter(year==2007) %>% summarize(pop = sum(pop)/(10^9))
```

*** 

### Part 2

**Write the code that reproduces these plots exactly**.  

**2A.** 

```{r, echo=FALSE, eval=TRUE}
ggplot(gapminder %>% filter(continent == 'Europe', 
                            year==2007),
       aes(x=pop,
           y=country)) +
  geom_col(fill='blue') +
  labs(title = 'Population sizes of European countries in 2007') +
  xlab('Population size') +
  ylab(NULL)
```

**2B.** 

```{r, echo=FALSE, eval=TRUE}
ggplot(gapminder,
       aes(x=continent, y=lifeExp)) +
  geom_violin(fill='darkblue', alpha=.5) +
  ylab('Life expectancy') +
  xlab('Continent') +
  labs(title='The geography of life expectancy')
```

**2C.** *(A more detailed view of the same topic)*

```{r, echo=FALSE, eval=TRUE}
ggplot(gapminder,
       aes(x=continent, y=lifeExp, color=year)) +
  geom_jitter(alpha=.7, size=2) +
  ylab('Life expectancy') +
  xlab('Continent') +
  labs(title='The geography of life expectancy')
```

**Hint:** The points on this plot have been *"jittered"* so that they do not fully overlap each other, and their transparency has been adjusted so that you can see where overlaps do occur. To replicate this, use `geom_jitter(alpha=.5)` instead of `geom_point()`. You can still adjust point size by adding `size` as an input within this `geom`.  


*** 

### Part 3

**Write the code that reproduces these plots exactly**.  

These questions will require two stages:  
- First, a new version of the dataset, grouped and summarized using `dplyr` verbs;  
- Second, a `ggplot` call.  

**3A.**  

```{r, echo=FALSE, eval=TRUE}
# First, summarize continent population size:
gap_conts <- 
  gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(pop = sum(pop)) %>% 
  mutate(pop = pop / 1000000000)

# Second, plot it:
ggplot(gap_conts, 
       aes(x=continent,
           y=pop)) + 
  geom_col(color='red', fill='pink') + 
  ylab('Population (in billions)') + 
  xlab(NULL) + 
  labs(title='Total population by continent in 2007')
```

**3B.** 

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# First, summarize continent population size:
gap_conts <- 
  gapminder %>% 
  group_by(year, continent) %>% 
  summarize(pop = sum(pop)) %>% 
    mutate(pop = pop / 1000000000)

# Second, plot it:
ggplot(gap_conts, 
       aes(x=year,
           color=continent,
           y=pop)) + 
  geom_line() + 
  xlab(NULL) + 
  ylab('Population (in billions)') + 
  labs(title='Population growth by continent, 1952 - 2007')
```


*** 

### Part 4

**Write the code that reproduces these plots exactly**.  



**Hint 1:** To log-transform an axis, add this line to your `ggplot()` call:

```{r, echo=TRUE, eval=FALSE}
# For x axis
+ scale_x_continuous(trans='log') 

# For y axis
+ scale_y_continuous(trans='log') 
```
  
  
**Hint 2:** To create a multi-panel plot that shows the same plot for different categories within the data, add this line to your `ggplot()` call:

```{r, echo=TRUE, eval=FALSE}
+ facet_wrap(~ variable)
```

Replace `variable` with the column that specifies the categories that you want represented in each "facet", or panel, of your plot. 
  
  
Here is an example with our class survey data: 

```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.width=8}
library(gsheet)
url <- 'https://docs.google.com/spreadsheets/d/1sW1X7KFuOIq9fddLD_otPWj3xJD24v0JFdDPh4vc1a4/edit?usp=sharing'
df <- gsheet2tbl(url)

ggplot(df, aes(x=height, y=shoe_size)) + 
  geom_point() + 
  facet_wrap(~gender)
```

Keep those hints as mind as you reproduce the following plots:  

**4A.** 

```{r, echo=FALSE, eval=TRUE}
ggplot(gapminder %>% filter(year %in% c(2007)),
       aes(x=lifeExp,
           y=gdpPercap,
           color=continent)) +
  geom_point() +
  scale_y_continuous(trans='log') + 
  xlab('Life expectancy in 2007') + 
  ylab('GPD per Capita (log-transformed)')
```

**4B.** 

```{r, echo=FALSE, eval=TRUE, fig.height=6, fig.width=8}
ggplot(gapminder %>% filter(year %in% c(1952, 1972, 1992, 2007)),
       aes(x=lifeExp,
           y=gdpPercap,
           size=pop,
           color=continent)) +
  geom_point() +
  scale_y_continuous(trans='log') +
  facet_wrap(~year) + 
  xlab('Life expectancy') + 
  ylab('GPD per Capita (log-transformed)')
```

